Executive Constraints and Civil Conflict Onset

### 2020-05-01 paper.Rmd
Manuscript edits

The $\delta$ and $\kappa$ parameters correspond to the slope and
intercept in the linear predictor for the presence of a legislature,
*v2lgbicam*. The effect of their inclusion can be seen in figure
\ref{fig:bicam}. The latent factor, $\theta$, follows the
directionality of the manifest variables, greater values indicate a
more constrained executive and lower values vice versa. The density
plots indicate that on average, for country-years where there is no
legislature, *i.e.* $v2lgbicam$ is zero, the latent factor is lower,
although still conditional on the values for the remaining manifest
variables. Where the legislature is present, $\theta$ is distributed
according to the $Normal(0, 1)$ prior.

@@

The final portion of the measurement model is the measurement error
adjustment. As described in the research design section, in order to
incorporate the posterior uncertainty from the V-Dem IRT aggregation,
the observed values for the manifest variables, $Z$, were assumed to
be drawn from normal distributions where the means were the unobserved
true values, $Z^{\ast}$, and the standard deviations the observed
posterior standard deviations as reported by V-Dem. The practical
effect is that for a given observation where the reported error is
larger, the uncertainty in the $Z^{\ast}_{ij}$ estimate increases and
the mean is drawn towards the regression line in the factor
analysis. In other words, the model is more skeptical of outliers in
the manifest variables. This posterior shrinkage can be seen in figure
\ref{fig:fa-line}. Figure \ref{fig:err} meanwhile plots the range of
the upper and low bounds of the $80$ percent posterior quantile
intervals for an example manifest variable in $Z^{\ast}$ against the
observed V-Dem standard deviations as well as the difference between
the estimated posterior mean and the observed mean.

@@

In total there are `r nrow(final.df)` country-year estimates of
$\theta$, comprised of `r n_distinct(final.df$country_name)` unique
countries and scaled from approximately $-2$ to $2$. Since the
complete time series for every single country can not be easily
visualised, an exploratory overview on the performance of the
executive constraints measure is provided in the form of several
different plots.  The first, figure \ref{fig:theta}, is a restricted
time series plot consisting of only China, Brazil, Nigeria, Norway,
and Russia. The countries were chosen to represent a diverse set of
geographic regions as well as political contexts. From the plot,
China, being a single party authoritarian state, is consistently
ranked the worst in executive constraints. The decline for Russia
largely follows the ascent of Vladimir Putin to the presidency at the
end of 1999 and his gradual consolidation of power in the decades that
followed.[^russia] For both Nigeria and Brazil, the measure captures
periods of military dictatorship such as the era of military rule in
Brazil following the 1964 coup d'etat and the period following Johnson
Aguiyi-Ironsi's seizure of power in the 1966 coup d'etat in Nigeria.
The last country, Norway, is an example of a consolidated polyarchy
for the time period under consideration with consistently strong
constraints against the executive office.

@@

Figure \ref{fig:reg} plots the slope coefficients on the log-odds
scale from the logistic regression linking the latent construct for
executive constraints to the outcome, civil conflict onset. Credible
intervals are summarised the same as previous plots, 80 percent
quantile intervals, and exact values are reported in table
\ref{tab:reg} together with the posterior probabilities that each
$\beta$ estimate is greater than zero and less than zero. Recall that
the variables are standardized to have mean 0 and standard deviation
1, with the exception of the ordinal/binary variables Relevant Group
Count, Neighbouring Conflict, Ongoing Major Conflict, and
Independence.

@@

This means that for the $\beta$ parameter for executive constraints
there is a `r as.numeric(beta_s[1, 2]) * 100` percent posterior
probability that the parameter is less than zero and a 80 percent
posterior probability that an increase of one standard deviation
decreases the log-odds of conflict onset between $`r beta_s[1, 1]`$. A
one standard deviation increase in this context corresponds roughly to
the difference between the level of executive constraints for Zambia
in 1967 under the administration of Kenneth Kaunda right before the
banning of opposition parties following the 1968 election, with score
of approximately 0, and Greece in 1976 shortly after the end of
military rule and establishment of a democratic republic, with a score
of approximately 1.

@@

Population density, mean elevation, number of political relevant
ethnic groups, proportion of the population facing state
discrimination, presence of neighbouring conflict, and the dummy
variable for independence year all increase the risk of civil conflict
onset. A one unit increase in population density is similar to going
from Cambodia in 1987 with a population of $`r cmb$pop`$ million and
an area of $`r cmb$area_sqkm`$ square kilometers to India in 1976 with
a population of $`r india$pop`$ million and an area of
$`r india$area_sqkm`$ square kilometers. A one unit increase for mean
elevation roughly matches the difference between Russia, $`r russia`$
meters, and Georgia, $`r georgia`$ meters. And, a one unit increase
for proportion discrimination is roughly equivalent to Thailand from
2002 until present with $`r thai`$ percent and Bhutan from 1989 with
$`r bhutan`$ percent.

Finally, the GDP per capita and executive constraints interaction term
and the dummy variable for ongoing high intensity conflict decrease
the risk of conflict onset. This latter result, although surprising,
may be due to the subsuming nature of high intensity civil wars. In
the face of an all-encompassing civil war, there may be a plethora of
factions with divergent interests active; however, their political
goals and interests are structured according to the current
incompatibility between the dominant political actors. Thus, there may
be new armed groups emerging who are willing to challenge to the
state, but that emergent may not necessarily be coded as a new civil
conflict. That said, it should be mentioned that the there is a much
higher degree of uncertainty in the estimate with the 80 percent
posterior credible interval ranging from -0.558 to 0.064.

@@

To interpret the results for the interaction term between the slopes
for executive constraints and GDP per Capita I calculate the marginal
effect of $\theta$, executive constraints, on the predicted
probability of civil conflict onset for different levels of GDP per
capita while holding constant the other covariates.  In this case, I
took the mean of GDP per capita growth, population density, mean
elevation, politically relevant ethnic group count, and peace years
while setting the dummy indicators for independence, ongoing conflict,
and neighbouring conflict and the random intercepts for year and
country to 0. The levels of GDP per capita were selected based on the
10\%, 50\%, and 90\% quantiles of the standardized log transformed
values. These levels roughly correspond to the GDP per capita of
Rwanda in 2007 at $`r gdp_rwa`$, the Philippines in 2008 at $`r gdp_phl`$,
and Portugal in 2009 at $`r gdp_po`$. In the dataset there are 60
conflict episode onsets below the first level, $152$ between the first
and second, $55$ between the second and third, and $4$ above the
third.

```{r fig.cap = "\\label{fig:int} Interaction between executive constraints and GDP per capita with 80\\% CIs"}
# Interaction plot
alpha <- extract(reg, "alpha")
theta_seq <- seq(-2.5, 2.5, .01)

state <- quantile(X[, "rgdpepc"], probs = c(0.1, .5, 0.9))

f <- function(x, y) {
    with(as.data.frame(X),
         alpha +
         beta[, 1] * x +
         beta[, 2] * y +
         beta[, 3] * mean(rgdpepc_gro) +
         beta[, 4] * mean(cinc) +
         beta[, 5] * mean(discrimpop) +
         beta[, 6] * mean(pop_density) +
         beta[, 7] * mean(meanelev) +
         beta[, 8] * mean(rlvt_groups_count) +
         beta[, 9] * 0 +
         beta[, 10] * 0 +
         beta[, 11] * 0 +
         beta[, 12] * x * y)
}

df <- lapply(state, function(y) {
    mclapply(theta_seq, function(point) {
        p_hat <- boot::inv.logit(f(point, y)) * 100
        post_summarise(p_hat, probs = c(0.1, 0.5, 0.9)) %>%
            mutate(theta = point,
                   y = y)
    }, mc.cores = 4) %>% bind_rows
}) %>% bind_rows

>>>>

The $\delta$ and $\kappa$ parameters correspond to the slope and
intercept in the linear predictor for the presence of a legislature,
*v2lgbicam*. The effect of their inclusion can be seen in figure
\ref{fig:bicam}. The latent factor, $\theta$, follows the
directionality of the manifest variables, greater values indicate a
more constrained executive and lower values vice versa. The density
plots indicate that on average, for country-years where there is no
legislature, *i.e.* $v2lgbicam$ is zero, the latent factor is lower,
although still conditional on the values for the remaining manifest
variables. Where the legislature is present, $\theta$ is distributed
according to the $\mathsf{Normal}(0, 1)$ prior.

@@

The final portion of the measurement model is the measurement error
adjustment. As described in the research design section, in order to
incorporate the posterior uncertainty from the V-Dem IRT aggregation,
the observed values for the manifest variables, $Z$, were assumed to
be drawn from normal distributions where the means were the unobserved
true values, $Z^{\ast}$, and the standard deviations the observed
posterior standard deviations as reported by V-Dem. The practical
effect is that for a given observation where the reported error is
larger, the uncertainty in the $Z^{\ast}_{ij}$ estimate increases and
the mean is drawn towards the regression line in the factor
analysis. In other words, the model is more skeptical of outliers in
the manifest variables. This posterior shrinkage can be seen in figure
\ref{fig:fa-line}. Figure \ref{fig:err} meanwhile plots the range of
the upper and low bounds of the $80$ percent posterior quantile
intervals for an example manifest variable in $Z^{\ast}$ against the
observed V-Dem standard deviations as well as the difference between
the estimated posterior mean and the observed mean in order to further
demonstrate this effect.

@@

In total there are `r nrow(final.df) %>% prettyNum(big.mark = ",")`
country-year estimates of $\theta$, comprised of `r
n_distinct(final.df$country_name)` unique countries and scaled from
approximately $-2$ to $2$. Since the complete time series for every
single country can not be easily visualised, an exploratory overview
on the performance of the executive constraints measure is provided in
the form of several different plots.  The first, figure
\ref{fig:theta}, is a restricted time series plot consisting of only
China, Brazil, Nigeria, Norway, and Russia. The countries were chosen
to represent a diverse set of geographic regions as well as political
contexts. From the plot, China, being a single party authoritarian
state, is consistently ranked the worst in executive constraints. The
decline for Russia largely follows the ascent of Vladimir Putin to the
presidency at the end of 1999 and his gradual consolidation of power
in the decades that followed.[^russia] For both Nigeria and Brazil,
the measure is able to distinguish between democratic rule and periods
of military dictatorship such as the era of military rule in Brazil
following the 1964 coup d'etat and the period following Johnson
Aguiyi-Ironsi's seizure of power in the 1966 coup d'etat in
Nigeria. The last country, Norway is an example of a consolidated
polyarchy for the time period under consideration with consistently
strong constraints against the executive office which are accurately
captured by the measure.

@@

Figure \ref{fig:reg} plots the slope coefficients from the logistic
regression linking the latent construct for executive constraints to
the outcome, civil conflict onset, on the log-odds scale. Credible
intervals are summarised the same as previous plots, 80 percent
quantile intervals, and exact values are reported in table
\ref{tab:reg} together with the posterior probabilities that each
$\beta$ estimate is greater than zero and less than zero. Recall that
the variables are standardized to have mean 0 and standard deviation
1, with the exception of the ordinal/binary variables Relevant Group
Count, Neighbouring Conflict, Ongoing Major Conflict, and
Independence.

@@

For the $\beta$ parameter for executive constraints there is a
`r as.numeric(beta_s[1, 2]) * 100` percent posterior probability that the
parameter is less than zero and a 80 percent posterior probability
that an increase of one standard deviation decreases the log-odds of
conflict onset between $`r beta_s[1, 1]`$. A one standard deviation
increase in this context corresponds roughly to the difference between
the level of executive constraints for Zambia in 1967 under the
administration of Kenneth Kaunda right before the banning of
opposition parties following the 1968 election, with a score of
approximately 0, and Greece in 1976 shortly after the end of military
rule and the establishment of a democratic republic with a score of
approximately 1.

@@

Population density, mean elevation, number of political relevant
ethnic groups, proportion of the population facing state
discrimination, presence of neighbouring conflict, and the dummy
variable for independence year all increase the risk of civil conflict
onset. A one unit increase in population density is similar to going
from Cambodia in 1987 with a population of $`r cmb$pop`$ million and
an area of $`r cmb$area_sqkm`$ square kilometers to India in 1976 with
a population of $`r india$pop`$ million and an area of
$`r india$area_sqkm`$ square kilometers. A one unit increase for mean
elevation roughly matches the difference between Russia at $`r russia`$
meters and Georgia at $`r georgia`$ meters. And, a one unit increase
for proportion of the population facing discrimination is roughly
equivalent to Thailand from 2002 until present with $`r thai`$ percent
and Bhutan from 1989 with $`r bhutan`$ percent.

Finally, the GDP per capita and executive constraints interaction term
and the dummy variable for ongoing high intensity conflict decrease
the risk of conflict onset. This latter result, although surprising,
may be due to the subsuming nature of high intensity civil wars. In
the face of an all-encompassing civil war, there may be a plethora of
active factions with divergent interests; however, their political
goals and interests are structured according to the current
incompatibility between the dominant political actors. Thus, there may
be new armed groups emerging who are willing to challenge to the
state, but that emergence may not necessarily be coded as a new civil
conflict. It should be noted, however, that there is a high degree of
uncertainty in the estimate with the 80 percent posterior credible
interval ranging from -0.558 to 0.064.

@@

To interpret the results for the interaction term between the slopes
for executive constraints and GDP per Capita I calculate the marginal
effect of $\theta$, executive constraints, on the predicted
probability of civil conflict onset for different levels of GDP per
capita while holding constant the other covariates.  In this case, I
take the mean of GDP per capita growth, CINC, the proportion of
discriminated population, population density, mean elevation, and the
number of politically relevant ethnic groups. I also set the dummy
indicators for independence, major ongoing conflict, and neighbouring
conflict and the random intercepts for year and country to zero. For
peace years I take the intercept for the average number of peace years
rounded to the nearest integer, which turns out to be 19. The levels
of GDP per capita were selected based on the 10\%, 50\%, and 90\%
quantiles of the standardized log transformed values. These levels
roughly correspond to the GDP per capita of Rwanda in 2007 at
$`r gdp_rwa`$ 2011 US dollars, the Philippines in 2008 at $`r gdp_phl`$
2011 US dollars, and Portugal in 2009 at $`r gdp_po`$ 2011 US
dollars. In the dataset there are 60 conflict episode onsets below the
first level, 152 between the first and second, 55 between the second
and third, and 4 above the third.

```{r fig.cap = "\\label{fig:int} Interaction between executive constraints and GDP per capita with 80\\% CIs"}
# Interaction plot
theta_seq <- seq(-2.5, 2.5, .01)
state <- quantile(X[, "rgdpepc"], probs = c(0.1, .5, 0.9))

marginal_data.df <- as.data.frame(X) %>%
    summarise_all(~mean(.)) %>%
    mutate(rlvt_groups_count = round(rlvt_groups_count),
           neighbour_conflict = 0,
           major_ongoing = 0,
           independence = 0)

peace_yr <- f[, mean(final.df$lpeace_yrs) %>% round]

fn <- function(x, y) {
    with(marginal_data.df,
         alpha +
         beta[, 1] * x +
         beta[, 2] * y +
         beta[, 3] * rgdpepc_gro +
         beta[, 4] * cinc +
         beta[, 5] * discrimpop +
         beta[, 6] * pop_density +
         beta[, 7] * meanelev +
         beta[, 8] * rlvt_groups_count +
         beta[, 9] * neighbour_conflict +
         beta[, 10] * major_ongoing +
         beta[, 11] * independence +
         beta[, 12] * x * y +
         peace_yr)
}

df <- lapply(state, function(y) {
    mclapply(theta_seq, function(point) {
        p_hat <- boot::inv.logit(fn(point, y)) * 100
        post_summarise(p_hat, probs = c(0.1, 0.5, 0.9)) %>%
            mutate(theta = point,
                   y = y)
    }, mc.cores = 4) %>% bind_rows
}) %>% bind_rows
